Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-20 HickernellMCQMC2024_QMCTutorial.tex
Updates from Overleaf

\vspace{2mm}
\noindent
\textbf{Combinatorial Optimization.}
For several decades, combinatorial optimization techniques have been employed to obtain good generating parameters for LD constructions. For example, in the case of a lattice rule which is fully determined by the generating vector \(\boldsymbol{h} \in \mathbb{Z}_n^d \coloneqq \{0, 1, \ldots, n-1\}^d\) given $n$ and $d$, 
%While good generating vectors are known for \(d \leq 2\) \cite{DicEtal22a,SloJoe94}, the same does not hold for \(d \geq 3\), 
one often resorts to finding good $\boldsymbol{h}$ via computer search to minimize a chosen FOM. An exhaustive search is infeasible due to the exponential growth of the search space with \(d\). Thus, the component-by-component (CBC) construction, introduced by Korobov \cite{kor63} and later revisited by Sloan and collaborators \cite{Slo02}, reduces the search space to size \(dn\) by constructing \(\boldsymbol{h}\) one component at a time while keeping the previous components fixed. This greedy procedure also applies to finding good generating matrices $\mathsf{C}_j$ for digital sequences.
%Further simplification is possible by restricting the search to Korobov lattices of the form \(\boldsymbol{h} = (1, h, h^2, \ldots, h^{d-1}) \in \mathbb{Z}^d\). 
For certain choices of function spaces, there exist fast component-by-component construction implementations using fast Fourier transforms (FFTs) as the primary tool. We refer to \cite{DicKuo04a,JoeKuo03,KuoJoe02b} and references therein for works on the CBC construction for lattice rules and digital sequences, \cite{NuyCoo06a,NuyCoo06b} for their fast implementations and \cite{LatNet} for a QMC software tool called LatNet Builder by L'Ecuyer to generate LD nodes which relies heavily upon the CBC construction method.

\vspace{2mm}
\noindent
\textbf{Global Optimization.} Global optimization of LD nodes is often challenging due to the non-convex nature of the FOM as a function of the nodes. Traditional global optimization methods, such as gradient descent or its variants, struggle in this setting becoming trapped in suboptimal local minima and failing to identify the global optimum configuration of nodes.

One of the first works to globally optimize LD nodes, albeit a heuristic approach, was by Winker and Fang \cite{WinFan97b} employing the threshold accepting optimization algorithm to directly optimize the uniformity of the LD nodes with respect to the discrepancy FOM.

More recently, Rusch, Kirk, Bronstein, Lemieux and Rus \cite{ruschkirk24} developed the first machine-learning assisted optimization tool for LD node optimization in the Message-Passing Monte Carlo (MPMC) method. MPMC implements tools from geometric deep learning imposing a computational graph onto an initialized IID node set to transform the nodes to LD node set via a learned mapping, again guided by the minimization of a discrepancy-based FOM.

Lastly, we highlight the method proposed by Hinrichs and Oettershagen \cite{hinoet16}, which generates optimal LD nodes for a specific class of functions in two-dimensions. To address the non-convexity challenge, the authors decompose the global optimization problem into a large number of smaller, convex sub-problems.

\vspace{2mm}
\noindent
\textbf{Other Notable Methods.} 
There are several other attempts to construct optimal LD nodes, many of which are well summarized in the thesis of Cl\'{e}ment \cite{}. We highlight the subset selection method proposed in \cite{cle22} to choose from an $n$-element initialization set, the $k<n$ nodes which minimize a discrepancy FOM. A heuristic approach to this problem was later shown in \cite{cle24}. %Furthermore, a method to generate optimal star-discrepancy node sets for fixed $n$ and $d$ based on a non-linear programming approach was suggested in \cite{}.

It is commonplace in qMC constructions that equidistribution is required not just in the (possibly high dimensional) $d$-dimensional space, but also for all or specific low-dimensional projections; see Section \ref{sec:coordwts}. MatBuilder \cite{paulin2022} is a software tool developed precisely for this task; to generate digital sequence LD nodes when low-dimensional uniformity is critical. Uniformity constraints related to digital sequence stratification across boxes of the form \eqref{eq:element_box} are expressed as integer linear programming (ILP) problems and subsequently solved, while optimizing an objective function relating to the ILP reformulation, to determine the next columns in the generator matrices for the digital sequence.

@@

\section{Reformulating Our Problem} \label{sec:reformulate}
\subsection{Variable Transformations}
\subsection{Variation Reduction}
\subsection{Multilevel Methods}

@@

\section*{Appendix A}
The integral $I_{\cos}(d) := \int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r$ and its relative, the integral $I_{\sin}(d) := \int_{0}^{\infty} \sin(r) \exp(-r^2) \, r^{d-1} \rmd r$, may be evaluated iteratively for $d = 1, 2, \ldots$.  Note that using integration by parts:
\begin{align*}
I_{\cos}(d)
	&=  \left .   \cos(r) \exp(-r^2) \, \frac{r^{d}}{d}  \right \rvert_0^{\infty} \\
	& \qquad \qquad - \int_{0}^{\infty} [\cos(r) (-2r )\exp(-r^2) -\sin(r)  \exp(-r^2)] \,  \frac{r^{d}}{d} \rmd r \\
	&= \frac{2 I_{\cos}(d+2) + I_{\sin}(d+1)}{d} \\
	I_{\sin}(d) &=  \left .   \sin(r) \exp(-r^2) \, \frac{r^{d}}{d}  \right \rvert_0^{\infty} \\
	&\qquad \qquad - \int_{0}^{\infty} [\sin(r) (-2r )\exp(-r^2) + \cos(r)  \exp(-r^2)] \,  \frac{r^{d}}{d} \rmd r \\
&= \frac{2 I_{\sin}(d+2) - I_{\cos}(d+1)}{d}
\end{align*}
This implies that
\begin{align*}
	I_{\cos}(d+2) & = \frac{d I_{\cos}(d) - I_{\sin}(d+1)}{2},   \\
	I_{\sin}(d+2) &=   \frac{d I_{\sin}(d) + I_{\cos}(d+1)}{2},
\end{align*}

>>>>

For several decades, combinatorial optimization techniques have been employed to obtain good generating parameters for LD constructions. For example, in the case of a lattice rule which is fully determined by the generating vector \(\boldsymbol{h} \in \mathbb{Z}_n^d \coloneqq \{0, 1, \ldots, n-1\}^d\) given $n$ and $d$, 
%While good generating vectors are known for \(d \leq 2\) \cite{DicEtal22a,SloJoe94}, the same does not hold for \(d \geq 3\), 
one often resorts to finding good $\boldsymbol{h}$ via computer search to minimize a chosen FOM. An exhaustive search is infeasible due to the exponential growth of the search space with \(d\). Thus, the component-by-component (CBC) construction, introduced by Korobov \cite{kor63} and later revisited by Sloan and collaborators \cite{Slo02}, reduces the search space to size \(dn\) by constructing \(\boldsymbol{h}\) one component at a time while keeping the previous components fixed. This greedy procedure also applies to finding good generating matrices $\mathsf{C}_j$ for digital sequences.
%Further simplification is possible by restricting the search to Korobov lattices of the form \(\boldsymbol{h} = (1, h, h^2, \ldots, h^{d-1}) \in \mathbb{Z}^d\). 
For certain choices of function spaces, there exist fast component-by-component construction implementations using fast Fourier transforms (FFTs) as the primary tool. We refer to \cite{DicKuo04a,JoeKuo03,KuoJoe02b} and references therein for works on the CBC construction for lattice rules and digital sequences, \cite{NuyCoo06a,NuyCoo06b} for their fast implementations and \cite{LatNet} for a QMC software tool called LatNet Builder by L'Ecuyer to generate LD nodes which relies heavily upon the CBC construction method.

It is commonplace in qMC constructions that equidistribution is required not just in the (possibly high dimensional) $d$-dimensional space, but also for all or specific low-dimensional projections; see Section \ref{sec:coordwts}. MatBuilder \cite{paulin2022} is a software tool developed precisely for this task; to generate digital sequence LD nodes when low-dimensional uniformity is critical. Uniformity constraints related to digital sequence stratification across boxes of the form \eqref{eq:element_box} are expressed as integer linear programming (ILP) problems and subsequently solved to optimize an objective function relating to the ILP reformulation. The solutions then determine the next columns in the generator matrices for the digital sequence.

As discussed above, the Halton sequence has the benefit of not having a preferred number of points for its construction. It does however possess the drawback of exhibiting poor uniformity in the lower dimensional projections when $d$ is moderate to high due to intrinsic number theoretic correlation between the coordinates \cite{kirklem24}. One popular strategy to circumvent this poor behavior, dating back to Faure \cite{fau78} and Braaten and Weller \cite{brawell79}, is to randomly or deterministically permute the digits of the base $b_j$ expansions $j = 1,2\ldots,d$ to break the observed correlation between the components. To do so, we first define the generalized van der Corput sequence by obtaining a sequence $\Sigma = (\sigma_r)_{r\geq0}$ of permutations of $\{0,\ldots,b-1\}$ where the $i^{th}$ term of the one-dimensional sequence is defined as 
\begin{multline}\label{eq:gen_vdc}
    \phi_b^{\Sigma}(i_0 + i_1b + i_2 b^2 + \cdots ) := \sigma_0(i_0) b^{-1} + \sigma_1(i_1) b^{-2} + \sigma_2(i_2) b^{-3} + \cdots \in [0,1)
	\\
	 \text{where } i_0, \sigma_0(i_0), i_1, \sigma_1(i_1), \ldots \in \{0,\ldots b-1\}.
\end{multline}
The generalized Halton sequence is subsequently defined by choosing $d$ sequences of permutations $\Sigma_j = (\sigma_{j,r})_{r\geq0}, j = 1, 2,\ldots,d$ and setting
$$\bsx_i^{\text{G-Hal}} = \left( \phi^{\Sigma_1}_{b_1}(i), \ldots, \phi^{\Sigma_d}_{b_d}(i)\right), \quad i= 0,1,\ldots,$$
There exist many attempts to optimize the permutations defining the generalized sequence, most of which involve so-called linear digit scramblings. We point the reader towards \cite{} and the excellent survey of the work in this direction (until 2008) by Faure and Lemieux \cite{}.

So far, we have discussed the attempts of researchers to optimize specific classes of LD constructions; lattices and digital sequences, and Halton sequences.  There also exist several efforts, some very recently, to generate novel LD constructions which don't follow any number theoretic or any other underlying structure, but none-the-less, possess a high level of uniformity. These methods are often global optimization methods, which is often challenging due to the non-convex nature of the FOM as a function of the nodes. Traditional global optimization methods, such as gradient descent or its variants, struggle in this setting becoming trapped in suboptimal local minima and failing to identify the global optimum configuration of nodes. One of the first works to globally optimize LD nodes, albeit a heuristic approach, was by Winker and Fang \cite{WinFan97b} employing the threshold accepting optimization algorithm to directly optimize the uniformity of the LD nodes with respect to the discrepancy FOM. More recently, Rusch, Kirk, Bronstein, Lemieux and Rus \cite{ruschkirk24} developed the first machine-learning assisted optimization tool for LD node optimization in the Message-Passing Monte Carlo (MPMC) method. MPMC implements tools from geometric deep learning imposing a computational graph onto an initialized IID node set to transform the nodes to LD node set via a learned mapping, again guided by the minimization of a discrepancy-based FOM. We also highlight the method proposed by Hinrichs and Oettershagen \cite{hinoet16}, which generates optimal LD nodes for a specific class of functions in two-dimensions. To address the non-convexity challenge, the authors decompose the global optimization of a specific FOM into a large number of smaller, convex sub-problems which are solved individually. There are several other attempts to construct optimal LD nodes, many of which are well summarized in the thesis of Cl\'{e}ment \cite{}. We highlight the subset selection method proposed in \cite{cle22} to choose from an $n$-element initialization set, the $k<n$ nodes which minimize a discrepancy FOM. A heuristic approach to this problem was later shown in \cite{cle24}. Furthermore, a method to generate optimal star-discrepancy node sets for fixed $n$ and $d$ based on a non-linear programming approach was suggested in \cite{}.

@@

\section{Reformulating Our Problem} \label{sec:reformulate}
\subsection{Variable Transformations}
\subsection{Variation Reduction}
\subsection{Multilevel Methods} \label{sec:multi}
The cost of computing the sample mean defined in \eqref{eq:sample_mean} is typically $\mathcal{O}(dn)$ as the dimension, $d$, and the sample size, $n$, tend to infinity.  Some applications have $d$ either huge or infinite.  On example is in quantitative finance, where $d$ denotes the number of times that the asset price is monitored over the life of a financial derivative times the number of assets involved.  However, there is a way to reduce the cost of approximating $\mu$ by employing multilevel methods \cite{}.

Briefly, let $\{Y_l := f_l(\bsX_{l})\}_{l=1}^L$ be a sequence of approximations to $Y = f(\bsX)$, where $\bsX_l \sim \mathcal{U}[0,1)^{d_l}$.  Here we assume that $d_1 < \cdots < d_L$, and that the cost of evaluating $f_l(\bsx_{l})$ is $\mathcal{O}(d)$.  Also, $d_L = d$, $Y_{d_L} = T$, and $f_L = f$. Then
\begin{align*}
    \mu & := \mathbb{E}(Y) = \mathbb{E}\bigl[f(\bsX)\bigr] 
    = \mathbb{E}(Y_1) + \mathbb{E}(Y_2 - Y_1) + \cdots + \mathbb{E}(Y_{d_{L-1}} - Y_d) \\
    & = \int_{[0,1)^{d_1}} f_1(\bsx^{(1)}) \, \mathrm{d} \bsx^{(1)} +  \int_{[0,1)^{d_2}} \bigl[f_2(\bsx^{(2)}) - f_1(\bsx^{(1)})\bigr] \, \mathrm{d} \bsx^{(2)} + \cdots \\
    & \qquad \qquad + \int_{[0,1)^{d_L}} \bigl[f_L(\bsx^{(L)}) - f_{L-1}(\bsx^{(L-1)}) \bigr] \, \mathrm{d} \bsx^{(L)}
    \intertext{where}
    \bsx^{(1)} & = (x_1, \dots, x_{d_1}), \quad \bsx^{(2)} = (x_1, \dots, x_{d_2}), \ldots, \bsx^{(L)} = (x_1, \dots, x_{d_L}).
\end{align*}
This series of integrals may be approximated by a series of sample means, namely
\begin{multline*}
    \hat{\mu}^{\text{ML}}_{\bsn,\bsd}  := \frac{1}{n_1}  \sum_{i=0}^{n_1-1} f_1(\bsx^{(1)}_i)  +  \frac{1}{n_2}  \sum_{i=0}^{n_2-1}  \bigl[f_2(\bsx^{(2)}_i) - f_1(\bsx^{(1)}_i)\bigr] + \cdots \\
     + \frac{1}{n_L}  \sum_{i=0}^{n_L-1}  \bigl[f_L(\bsx_i^{(L)}) - f_{L-1}(\bsx_i^{(L-1)}) \bigr].
\end{multline*}
The computational cost of this approximation is $\mathcal{O}(d_1n_1 + \cdots + d_Ln_L)$.

A worst case error analysis as described in Section \ref{sec:discrepancy} implies that
\begin{multline*}
    \bigl \lvert \mu - \hat{\mu}^{\text{ML}}_{\bsn,\bsd} \bigr \rvert \le 
    \text{discrepancy}\bigl(\{\bsx_i^{(1)}\}_{i=0}^{n_1-1}, K^{(1)} \bigr) \text{variation}(f^{(1)}, K^{(1)}) \\
    + \text{discrepancy}\bigl(\{\bsx_i^{(2)}\}_{i=0}^{n_2-1}, K^{(2)} \bigr) \text{variation}(f^{(2)} - f^{(1)}, K^{(2)}) + \cdots \\
    + \text{discrepancy}\bigl(\{\bsx_i^{(L)}\}_{i=0}^{n_L-1}, K^{(L)} \bigr) \text{variation}(f^{(L)} - f^{(L-1)}, K^{(L)}),
\end{multline*}
where $K^{(l)}$ is the reproducing kernel corresponding to the relevant  Hilbert space of functions of $d_l$ variables.

@@

\end{document}
\section*{Appendix A}
The integral $I_{\cos}(d) := \int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r$ and its relative, the integral $I_{\sin}(d) := \int_{0}^{\infty} \sin(r) \exp(-r^2) \, r^{d-1} \rmd r$, may be evaluated iteratively for $d = 1, 2, \ldots$.  Note that using integration by parts:
\begin{align*}
I_{\cos}(d)
	&=  \left .   \cos(r) \exp(-r^2) \, \frac{r^{d}}{d}  \right \rvert_0^{\infty} \\
	& \qquad \qquad - \int_{0}^{\infty} [\cos(r) (-2r )\exp(-r^2) -\sin(r)  \exp(-r^2)] \,  \frac{r^{d}}{d} \rmd r \\
	&= \frac{2 I_{\cos}(d+2) + I_{\sin}(d+1)}{d} \\
	I_{\sin}(d) &=  \left .   \sin(r) \exp(-r^2) \, \frac{r^{d}}{d}  \right \rvert_0^{\infty} \\
	&\qquad \qquad - \int_{0}^{\infty} [\sin(r) (-2r )\exp(-r^2) + \cos(r)  \exp(-r^2)] \,  \frac{r^{d}}{d} \rmd r \\
&= \frac{2 I_{\sin}(d+2) - I_{\cos}(d+1)}{d}
\end{align*}
This implies that
\begin{align*}
	I_{\cos}(d+2) & = \frac{d I_{\cos}(d) - I_{\sin}(d+1)}{2},   \\
	I_{\sin}(d+2) &=   \frac{d I_{\sin}(d) + I_{\cos}(d+1)}{2},
\end{align*}
